---
title: "Differential gene expresion analysis of tissue vs. organoids"
output: html_notebook
---

```{r include=F}
library('tidyverse')
library('edgeR')
library('ComplexHeatmap')
library('VennDiagram')
library('Vennerable')
```

Specify count data file path, read it, and format
```{r}
countsFp = '../data/rnaseq_sample_counts.txt'
counts_df = read_delim(countsFp, delim='\t', skip=1, show_col_types=F) %>% 
  select(-c(2:6)) %>% 
  janitor::clean_names()

colnames(counts_df) = gsub("work_las_nvalenzu_lab_collaboration_read_mapping_star_", "", colnames(counts_df)) %>%
  gsub("aligned_sorted_by_coord_out_bam", "", .) %>% 
  gsub("_$", '', .)
```

Remove samples from data set
```{r}
counts_df = counts_df[, !grepl("_1$|_2$|ovary", colnames(counts_df))]
```

Extract spike-in counts
```{r}
ercc_df = counts_df %>%
  filter(grepl('^ERCC-', .$geneid))

counts_df = counts_df %>%
  filter(!grepl('^ERCC-', .$geneid))
```

Create table with sample groups
```{r}
samplemeta = tibble(sample=colnames(counts_df[, -1])) %>%
  mutate(group=str_extract(sample, pattern="_[_a-z]+") %>%
           gsub("^_|_$", "", .)) %>%
  mutate(ind=str_extract(sample, pattern="b[0-9]+"))
```

Calculate normalization factors from spike-ins
```{r}
normfactors = calcNormFactors(column_to_rownames(ercc_df, var='geneid'))
```

Create DGEList from counts and sample groups
```{r}
counts = DGEList(counts=column_to_rownames(counts_df, var='geneid'), 
                 norm.factors=normfactors,
                 group=samplemeta$group)
```

Plot raw and normalized counts for comparison
```{r include=F, message=F}
logcounts = cpm(counts, log=T, prior.count=0.5) %>%
  as.data.frame() %>%
  rownames_to_column(var='gene')
cpm_p = list()
for(i in colnames(counts)){
  cpm_p[[i]] = ggplot(counts_df, aes_string(x=i)) + 
    geom_histogram() #+
    #xlab(i)
  
  cpm_p[[paste0(i, '_log')]] = ggplot(logcounts, aes_string(x=i)) + 
    geom_histogram() #+
    #xlab(i)
}
#pdf('../results/density_raw_and_logCPM.pdf')
ggpubr::ggarrange(plotlist=cpm_p, ncol=2,nrow=2)
#dev.off()
```

Generate MDS for all samples
```{r}
# Filter lowly expressed genes
keep = filterByExpr(counts, group=counts$samples$group)
countsFilter = counts[keep, ]
mds = plotMDS(countsFilter, plot=F)
mds_df = tibble(sample=colnames(countsFilter),
                group=samplemeta$group,
                dog=samplemeta$ind,
                dim1=mds$x,
                dim2=mds$y)

#pdf('../results/MDS_expr_profiles.pdf', width=9, height=7)
ggplot(mds_df, aes(x=dim1, y=dim2, color=group, shape=dog, label=group)) +
  geom_point(size=4) +
  ggrepel::geom_text_repel(color='black') +
  xlab(paste0("Dim1 - Explained var=", round(mds$var.explained[1] * 100, 1), '%')) +
  ylab(paste0("Dim2 - Explained var=", round(mds$var.explained[2] * 100, 1), '%')) +
  ggtitle('Multidimensional scaling plot of log gene expression') +
  khroma::scale_color_discreterainbow()
#dev.off()
```

Detect DE genes
```{r}
tissues = c('bladder', 'endometrium', 'kidney', 'liver', 'lung', 'pancreas')
pval_h = list()
all_tests = list()
for(i in tissues){
  # Subset DGEList to only relevant samples
  counts_tmp = counts[, grep(i, colnames(counts))]
  
  # Create design matrix
  # Option 1: Simple model to test for differences between tissue vs. organoid
  design = model.matrix(~samplemeta$group[grep(i, samplemeta$group)])
  colnames(design)[2] = c(as.vector(na.omit(str_extract(colnames(design), '[a-z]+_[a-z]+'))))
  
  # Option 2: Test for differences between tissue vs. organoids, adjusting for differences between samples (Probably more accurate model)
  # design = model.matrix(~samplemeta$group[grep(i, samplemeta$group)] + samplemeta$ind[grep(i, samplemeta$group)])
  # colnames(design)[c(2,3)] = c(as.vector(na.omit(str_extract(colnames(design), '[a-z]+_[a-z]+'))),
  #                              as.vector(na.omit(str_extract(colnames(design), 'b[8186]+')))
  #                              )
  
  rownames(design) = colnames(counts_tmp)
  
  keep = filterByExpr(counts_tmp, design=design)
  counts_tmp = counts_tmp[keep, ]
  
  counts_tmp = estimateDisp(counts_tmp, design, robust=T)
  
  fit = glmQLFit(counts_tmp, design)
  
  qlft = glmQLFTest(fit)
  
  # Plot histogram of nominal pvalues
  pval_h[[i]] = ggplot(qlft$table) + 
    geom_histogram(aes(x=PValue)) +
    ggtitle(i)
  
  # Use two methods for multiple sample comparison p value correction
  topDE_Holm = topTags(qlft, n=40000, adjust.method='holm')
  topDE_BH = topTags(qlft, n=40000, adjust.method='BH')
  
  all_tests[[i]] = list(BH=topDE_BH, Holm=topDE_Holm)
  
  de_list = list(
    DEGenes_BH = topDE_BH$table %>%
      rownames_to_column('gene_symbol') %>%
      mutate(abslogFC=abs(.$logFC)) %>%
      arrange(FDR, desc(abslogFC)) %>%
      filter(FDR < 0.05) %>%
      select(-abslogFC),
    DEGenes_Holm = topDE_Holm$table %>%
      rownames_to_column('gene_symbol') %>%
      mutate(abslogFC=abs(.$logFC)) %>%
      arrange(FWER, desc(abslogFC)) %>%
      filter(FWER < 0.05) %>%
      select(-abslogFC),
    Non_DEGenes_BH = topDE_BH$table %>%
      rownames_to_column('gene_symbol') %>%
      arrange(desc(logCPM)) %>%
      filter(FDR >= 0.05) %>%
      filter(logCPM > quantile(.$logCPM, 0.5)),
    Non_DEGenes_Holm = topDE_Holm$table %>%
      rownames_to_column('gene_symbol') %>%
      arrange(desc(logCPM)) %>%
      filter(FWER >= 0.05) %>%
      filter(logCPM > quantile(.$logCPM, 0.5))
  )
  
  #openxlsx::write.xlsx(x=de_list, paste0(file='../results/top_DE_genes_tissue_vs_organoid_', i, '.xlsx'))
}

#pdf('../results/p_value_histograms.pdf')
ggpubr::ggarrange(plotlist=pval_h, nrow=n2mfrow(length(pval_h))[1], ncol=n2mfrow(length(pval_h))[2])
#dev.off()
```

Generate heatmaps
```{r}
fps = list.files('../results/tissue_vs_organoid/', pattern='xlsx', full.names=T)
hm_BH = list()
hm_Holm = list()
for(i in tissues){
  
  degenes_BH = readxl::read_excel(grep(i, fps, value=T), sheet=1) %>%
    filter(logFC <= quantile(.$logFC, 0.01) | logFC >= quantile(.$logFC, 0.99))
  
  degenes_Holm = readxl::read_excel(grep(i, fps, value=T), sheet=2) %>%
    filter(logFC <= quantile(.$logFC, 0.01) | logFC >= quantile(.$logFC, 0.99))
  
  if(nrow(degenes_BH) > 0){
    mtx_BH = cpm(countsFilter$counts, log=T, prior.count=0.5) %>%
      as.data.frame() %>%
      select(grep(i, colnames(.))) %>%
      rownames_to_column(var='gene_symbol') %>%
      right_join(., degenes_BH[, c(1,2,6)], by='gene_symbol') %>%
      arrange(logFC) %>%
      filter(FDR < 0.01) %>%
      select(-c(logFC, FDR)) %>%
      column_to_rownames(var='gene_symbol') %>%
      as.matrix()
    colnames(mtx_BH) = gsub(paste0(i, '_|anoid|sue'), '', colnames(mtx_BH))
    hm_BH[[i]] = Heatmap(scale(mtx_BH, center=F), cluster_rows=F,
                         column_title=paste0(i, ' DE genes, Benjamini-Hochberg/FDR<0.01\nlogFC<=qt0.01 or logFC>=qt0.99'),
                         name='norm_expr')
  }
  
  if(nrow(degenes_Holm) > 0){
    mtx_Holm = cpm(countsFilter$counts, log=T, prior.count=0.5) %>%
      as.data.frame() %>%
      select(grep(i, colnames(.))) %>%
      rownames_to_column(var='gene_symbol') %>%
      right_join(., degenes_Holm[, c(1,2,6)], by='gene_symbol') %>%
      arrange(logFC) %>%
      filter(FWER < 0.05) %>%
      select(-c(logFC, FWER)) %>%
      column_to_rownames(var='gene_symbol') %>%
      as.matrix()
    colnames(mtx_Holm) = gsub(paste0(i, '_|anoid|sue'), '', colnames(mtx_Holm))
    hm_Holm[[i]] = Heatmap(scale(mtx_Holm, center=F), cluster_rows=F, 
                           column_title=paste0(i, ' DE genes, Holm/FWER<0.01\nlogFC<=qt0.01 or logFC>=qt0.99'),
                           name='norm_expr')
  }
  
}
```

Write heatmaps to file
```{r}
i = 'bladder'
#pdf(paste0('../results/DE_genes_tissue_vs_organoid_BHPadjust_', i, '.pdf'), height=24, pointsize=8)
print(hm_BH[[i]])
#dev.off()

i = 'endometrium'
#pdf(paste0('../results/DE_genes_tissue_vs_organoid_BHPadjust_', i, '.pdf'), height=28, pointsize=8)
print(hm_BH[[i]])
#dev.off()

i = 'kidney'
#pdf(paste0('../results/DE_genes_tissue_vs_organoid_BHPadjust_', i, '.pdf'), height=7, pointsize=8)
print(hm_BH[[i]])
#dev.off()

i = 'liver'
#pdf(paste0('../results/DE_genes_tissue_vs_organoid_BHPadjust_', i, '.pdf'), height=30, pointsize=8)
print(hm_BH[[i]])
#dev.off()

i = 'lung'
#pdf(paste0('../results/DE_genes_tissue_vs_organoid_BHPadjust_', i, '.pdf'), height=30, pointsize=8)
print(hm_BH[[i]])
#dev.off()

i = 'pancreas'
#pdf(paste0('../results/DE_genes_tissue_vs_organoid_BHPadjust_', i, '.pdf'), height=30, pointsize=8)
print(hm_BH[[i]])
#dev.off()
```

```{r}
i = 'bladder'
#pdf(paste0('../results/DE_genes_tissue_vs_organoid_HolmPadjust_', i, '.pdf'), height=7)
print(hm_Holm[[i]])
#dev.off()

i = 'endometrium'
#pdf(paste0('../results/DE_genes_tissue_vs_organoid_HolmPadjust_', i, '.pdf'), height=7)
print(hm_Holm[[i]])
#dev.off()

# i = 'kidney'
# pdf(paste0('../results/DE_genes_tissue_vs_organoid_HolmPadjust_', i, '.pdf'), height=7)
# print(hm_Holm[[i]])
# dev.off()

i = 'liver'
#pdf(paste0('../results/DE_genes_tissue_vs_organoid_HolmPadjust_', i, '.pdf'), height=7)
print(hm_Holm[[i]])
#dev.off()

i = 'lung'
#pdf(paste0('../results/DE_genes_tissue_vs_organoid_HolmPadjust_', i, '.pdf'), height=10)
print(hm_Holm[[i]])
#dev.off()

i = 'pancreas'
#pdf(paste0('../results/DE_genes_tissue_vs_organoid_HolmPadjust_', i, '.pdf'), height=30)
print(hm_Holm[[i]])
#dev.off()
```

Venn diagrams
```{r include=F}
fps = list.files('../results/tissue_vs_organoid/', pattern='xlsx', full.names=T)
venn_BH = list()
org_Up = list()
tis_Up = list()
for(i in tissues){
  
  degenes_BH = readxl::read_excel(grep(i, fps, value=T), sheet=1) %>%
    arrange(logFC)
  
  org_Up[[i]] = degenes_BH$gene_symbol[degenes_BH$logFC < 0 & degenes_BH$FDR < 0.01]
  tis_Up[[i]] = degenes_BH$gene_symbol[degenes_BH$logFC > 0 & degenes_BH$FDR < 0.01]
}

venn_BH[['organoid']] = venn.diagram(x=org_Up[-4], disable.logging=T, filename=NULL,
                                     cat.pos=c(0, 350, 216, 144, 50),
                                     main='DE Up genes in organoids compared to tissues - BH/FDR < 0.01',
                                     fill=khroma::color('discrete rainbow')(5))

venn_BH[['tissue']] = venn.diagram(x=tis_Up[-4], disable.logging=T, filename=NULL,
                                   cat.pos=c(0, 350, 216, 144, 50),
                                   main='DE Up genes in tissues compared to organoids - BH/FDR < 0.01', 
                                   fill=khroma::color('discrete rainbow')(5))

#pdf('../results/organoid_DE_up_genes.pdf')
grid.draw(venn_BH$organoid)
#dev.off()

v1 = Venn(org_Up)
#pdf('../results/organoid_DE_up_genes_6setVenn.pdf')
plot(v1, doWeights=FALSE)
#dev.off()

#pdf('../results/tissue_DE_up_genes.pdf')
grid.draw(venn_BH$tissue)
#dev.off()

v2 = Venn(tis_Up)
#pdf('../results/tissue_DE_up_genes_6setVenn.pdf')
plot(v2, doWeights=FALSE)
#dev.off()

# Read Diff gene expression results to sort intersections by logFC
degstats = tibble()
for(i in tissues){
  degstats = degstats %>%
    bind_rows(readxl::read_excel(grep(i, fps, value=T), sheet=1))
}
degstats = degstats %>%
    dplyr::select(gene_symbol, logCPM, logFC, FDR) %>%
    #mutate(abs_logFC=abs(logFC)) %>%
    #arrange(gene_symbol, desc(abs_logFC))
    arrange(gene_symbol, desc(logCPM)) %>%
    filter(logFC < 0 & FDR < 0.01) %>%
    dplyr::select(-FDR)
degstats = degstats[!duplicated(degstats$gene_symbol), ]

# Save combinations of intersections for organoids
vennInts = list()
for(i in 1:length(org_Up)){
  # Get table with all possible combinations of elements, selecting i number of elemets at a time
  combs = combn(1:length(org_Up), i)
  for(j in 1:ncol(combs)){
    # Get the complement to the intersection (will be substracted from result)
    comp = c(1:length(org_Up))[!(1:length(org_Up) %in% combs[, j])]
    # Get name of intersection
    nameint = paste0(sort(names(org_Up)[combs[, j]]), collapse='_')
    # Get intersection of all elements in combs[, j]
    vennInts[[nameint]] = Reduce(intersect, org_Up[combs[, j]])
    # Substract complement
    vennInts[[nameint]] = vennInts[[nameint]][!(vennInts[[nameint]] %in% unlist(org_Up[comp]))]
    vennInts[[nameint]] = degstats[degstats$gene_symbol %in% vennInts[[nameint]], ] %>%
      arrange(desc(logCPM))
      #dplyr::select(-abs_logFC)
    vennInts[[nameint]] = bind_rows(tibble(gene_symbol=nameint, logCPM=NA, logFC=NA), vennInts[[nameint]])
  }
}
names(vennInts) = seq(1, length(vennInts))
#openxlsx::write.xlsx(x=vennInts, file="../results/organoid_DE_up_genes_6setVenn_Intersections.xlsx")

# Read Diff gene expression results to sort intersections by logFC
degstats = tibble()
for(i in tissues){
  degstats = degstats %>%
    bind_rows(readxl::read_excel(grep(i, fps, value=T), sheet=1))
}
degstats = degstats %>%
    dplyr::select(gene_symbol, logCPM, logFC, FDR) %>%
    #mutate(abs_logFC=abs(logFC)) %>%
    #arrange(gene_symbol, desc(abs_logFC))
    arrange(gene_symbol, desc(logCPM)) %>%
    filter(logFC > 0 & FDR < 0.01) %>%
    dplyr::select(-FDR)
degstats = degstats[!duplicated(degstats$gene_symbol), ]

# Save combinations of intersections for tissues
vennInts = list()
for(i in 1:length(tis_Up)){
  # Get table with all possible combinations of elements, selecting i number of elemets at a time
  combs = combn(1:length(tis_Up), i)
  for(j in 1:ncol(combs)){
    # Get the complement to the intersection (will be substracted from result)
    comp = c(1:length(tis_Up))[!(1:length(tis_Up) %in% combs[, j])]
    # Get name of intersection
    nameint = paste0(sort(names(tis_Up)[combs[, j]]), collapse='_')
    # Get intersection of all elements in combs[, j]
    vennInts[[nameint]] = Reduce(intersect, tis_Up[combs[, j]])
    # Substract complement
    vennInts[[nameint]] = vennInts[[nameint]][!(vennInts[[nameint]] %in% unlist(tis_Up[comp]))]
    vennInts[[nameint]] = degstats[degstats$gene_symbol %in% vennInts[[nameint]], ] %>%
      arrange(desc(logCPM))
      #dplyr::select(-abs_logFC)
    vennInts[[nameint]] = bind_rows(tibble(gene_symbol=nameint, logCPM=NA, logFC=NA), vennInts[[nameint]])    
  }
}
names(vennInts) = seq(1, length(vennInts))
#openxlsx::write.xlsx(x=vennInts, file="../results/tissue_DE_up_genes_6setVenn_Intersections.xlsx")
```

Do MDS diagram of DE genes
```{r}
# Merge all DE genes from all comparisons
degenes = unique(c(unlist(org_Up), unlist(tis_Up)))

# Filter lowly expressed genes
keep = filterByExpr(counts, group=counts$samples$group)
countsFilterMDS = counts[keep, ]
countsFilterMDS = countsFilterMDS[rownames(countsFilterMDS) %in% degenes, ]

mdsDEGenes = plotMDS(countsFilterMDS, plot=F)
mdsDEGenes_df = tibble(sample=colnames(countsFilterMDS),
                group=samplemeta$group,
                dog=samplemeta$ind,
                dim1=mdsDEGenes$x,
                dim2=mdsDEGenes$y)

#pdf('../results/MDS_expr_profiles_OnlyDEGenes.pdf', width=9, height=7)
ggplot(mdsDEGenes_df, aes(x=dim1, y=dim2, color=group, shape=dog, label=group)) +
  geom_point(size=4) +
  ggrepel::geom_text_repel(color='black') +
  xlab(paste0("Dim1 - Explained var=", round(mdsDEGenes$var.explained[1] * 100, 1), '%')) +
  ylab(paste0("Dim2 - Explained var=", round(mdsDEGenes$var.explained[2] * 100, 1), '%')) +
  ggtitle('Multidimensional scaling plot of log gene expression') +
  khroma::scale_color_discreterainbow()
#dev.off()
```

Pairwise Venn diagrams
```{r include=F}
for(i in tissues){
  expr_subset = logcounts[, grep(i, colnames(logcounts), value=T)]
  tissue_genes = logcounts$gene[rowMeans(expr_subset[, grep(paste0(i, '_tissue'), colnames(expr_subset), value=T)]) > mean(as.matrix(expr_subset))]
  org_genes = logcounts$gene[rowMeans(expr_subset[, grep(paste0(i, '_org'), colnames(logcounts), value=T)]) > mean(as.matrix(expr_subset))]
  
  
  venn_tissue = venn.diagram(x=list(tissue_genes, y=org_genes), disable.logging=T, filename=NULL,
                            cat.pos=c(315, 45), category.names=c(paste0(i, '\ntissue'), paste0(i, '\norganoid')),
                            main=paste0('Genes with expression over mean', ' - ', i),
                            fill=khroma::color('discrete rainbow')(2),
                            euler.d=T, scaled=T
                            )
  
  pdf(paste0('../results/tissue_vs_organoid/', i, '_up_genes.pdf'))
  grid.draw(venn_tissue)
  dev.off()
}
```


